Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-06-04 paper/paper.qmd
dupe
edit

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).

@@

The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation. In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.

@@

## Grouping Dynamics Across the Four Phases of the Aurignacian

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

@@

One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which change from phase to phase.

The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation. In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.

A potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984]. The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian. It is possible that those sites with a greater than expected diversity index acted as aggregation sites for their groups or served a similar purpose. La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it functioned as a specifically inter-group aggregation site.

@@

@fig-div-plot shows divergence from the expected diversity index for the sample size. Another potential explanation for the observed variation in diversity is that it may indicate variation in social structure. Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012]. Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012]. It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985; @Schwendler_2012]. This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.

>>>>

@@

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.

@@

## Grouping Dynamics Across the Four Phases of the Aurignacian
Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

One consistent grouping dynamic throughout all four phases is the restricted/broader group divide, which may be influenced by variation in social structure. Despite the tendency to conceive of hunter-gatherers as homogeneous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012], and additional studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012]. It is therefore possible that, for some of the sites, the broader/restricted grouping indicates variation in social structure, with the broader groups having a greater variety of formalized social roles, and the restricted groups having fewer formalized social roles.

@@

One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], inclusing, in some cases, decorative objects [@Kelly_2024]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

Aggregation activities may be identified by a higher than expected diversity index. @fig-div-plot plots the expected Shannon-Weaver diversity index [@Frerebeau_2025] for a given sample size against the actual diversity index per site [@Kintigh_1984]. The plot identifies multiple sites with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian. Notably, most of these sites are not shown as having connections between groups in @fig-neighbor-net, except for La Ferrassie in the Proto-Aurignacian. This suggests a distinction between shorter-term information exchange (e.g. the highly interconnected sites in @fig-neighbor-net) and longer-term residential movement (e.g. the sites with higher than expected diversity in @fig-div-plot).

@@
